---
title: "Dada2 pipeline in R"
author: "Marko Suokas"
format: pdf
pdf-engine: lualatex
editor: visual
mainfont: Aptos
always_allow_html: yes
header-includes:
   \usepackage[dvipsnames]{xcolor}
   \definecolor{darkblue}{rgb}{0.0, 0.0, 0.55}
   \definecolor{maroon}{rgb}{0.5, 0.0, 0.0}
   \definecolor{ivory}{rgb}{1.0, 1.0, 0.94}
   \definecolor{indigo}{rgb}{0.29, 0.0, 0.51}
---

```{r, include = F}

# Code to adjust text size inside R code chunks, default is normal size
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n",
                                              x, "\n\n \\normalsize"), x)
})

# Load libraries
library(dada2);packageVersion("dada2")
library(mia)
library(knitr);packageVersion("knitr")
library(Biostrings);packageVersion("Biostrings")
library(DECIPHER);packageVersion("DECIPHER")
library(tidyverse);packageVersion("tidyverse")
library(kableExtra);packageVersion("kableExtra")
library(patchwork);packageVersion("patchwork")
```

#### Parameters

```{r parameters, warning = F, message = F, size = "small"}
# Path to reads
path <- "data/reads"

# Path to taxonomy reference
refdb <- "~/reference/silva_nr99_v138.1_train_set.fa.gz"

# Path to metadata
meta_file <- "data/metadata.tsv"

# Path to export directory
exportloc <- "results"

# Dada settings
truncation <- 245
phi <- FALSE

#Name of first column in metadata file
meta_1stcol <- "Sampleid"

#Create results directory
dir.create(exportloc)
```

#### Sample lists

```{r lists, warning=F, message=F, size="small"}
#List files in path
list.files(path)
#Filenames have format: SAMPLENAME_R1_001.fastq
fnFs <- sort(list.files(path, pattern = "_R1_001.fastq", full.names = T))
# Extract sample names, assuming pattern
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
#Filtered files will be placed in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
names(filtFs) <- sample.names
```

**Tip:** If you have numbered samples, use 0X format. Otherwise you have problems in sort order.

\newpage

#### Quality profile

```{r qplot, warning=F, message=F, size = "small", fig.dim=c(6.5,6)}
# Base quality plot in first 6 samples
plotQualityProfile(fnFs[1:6])
```

\newpage

### Filtering and trimming reads

```{r filtering, warning=F, message=F, eval = F, size = "small"}
#Filtered files will be placed in filtered/ subdirectory
out <- filterAndTrim(fnFs, filtFs, truncLen=truncation,
                     maxN=0, maxEE=2, truncQ=2,
                     compress=TRUE, multithread=8)
#Output is saved to rds file, so we don't have to recalculate, if we make changes
#If you are making changes to chunk, change eval = TRUE
saveRDS(out,"rds/out.rds")
```

```{r rds1, warning = F, message = F, size = "small"}
#read rds file
out <- readRDS("rds/out.rds")
```

**Considerations:** The standard parameters are starting points. If you want to speed up downstream computation, consider tightening `maxEE`. If too few reads are passing the filter, consider relaxing `maxEE`.

For ITS sequencing, it is usually undesirable to truncate reads to a fixed length due to the large length variation at that locus. You can omit in this case truncLen parameter.

#### Learn error rates

Step determinates error rate of dataset using *learnErrors* function.

```{r learnerrors, warning=F, message=F, size = "small", eval = F}
# Forward read error rate
errF <- learnErrors(filtFs, multithread=8)
# saverds
saveRDS(errF,"rds/errF.rds")
```

```{r rds2, warning = F, message = F, size = "small"}
errF <- readRDS("rds/errF.rds")
```

\newpage

#### Plot error profiles

```{r errorplot, warning=F, message=F, size = "small", fig.dim = c(6.5,6)}
# Plotting error rate profile for forward reads
plotErrors(errF, nominalQ=TRUE)
```

\newpage

#### Denoise data

```{r denoise, warning=F, message=F, eval = F, size = "small"}
#denoise command
dadaFs <- dada(filtFs, err=errF, multithread=8)
#save to rds file
saveRDS(dadaFs,"rds/dadaFs.rds")
```

```{r rds3, warning = F, message = F, size = "small"}

#read rds
dadaFs <- readRDS("rds/dadaFs.rds")
```

#### Create ASV table

```{r asvtable, warning=F, message=F, size = "small"}
seqtab <- makeSequenceTable(dadaFs)
# Dimensions of ASV table
dim(seqtab)
```

#### Remove chimeric variants

```{r chimeras, warning=F, message=F, size = "small"}
seqtab.nochim <- removeBimeraDenovo(seqtab, method = "consensus",
                                    multithread = T, verbose = T)
dim(seqtab.nochim)
```

\newpage

#### Summary

Table can help to pinpoint if large amount of the data is lost at some point.

```{r summary, warning=F, message=F, size = "small"}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), rowSums(seqtab.nochim),
               rowSums(seqtab.nochim != 0))

# If processing a single sample, replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("Input", "Filtered", "Denoised", "Nonchimeric", "Variants")
rownames(track) <- sample.names
kable(track, caption="Denoising summary") %>%
  kable_styling(latex_options = c("HOLD_position","striped")) %>%
  row_spec(0, background="indigo", color="ivory")
```

\newpage

#### Assign taxonomy

We use assignTaxonomy from dada2 library and Silva database to assign taxonomic information.

```{r taxonomy, warning = F, message = F, size = "small", eval = F}

# Assign taxonomy
taxa <- assignTaxonomy(colnames(seqtab.nochim), refFasta=refdb,
                       multithread=8, verbose=T)
# Add Species column
taxa <- data.frame(taxa) %>% mutate(Species = NA)
# Save rds
saveRDS(taxa, "rds/taxa.rds")
```

```{r, warning = F, message = F, size = "small"}

taxa <- readRDS("rds/taxa.rds")
```

#### Create TSE object

```{r tse, warning=FALSE, message=FALSE, size = "small"}

# Counts table is a matrix containing abundance information
counts <- t(seqtab.nochim)
rownames(counts) <- NULL

# Metadata
metafile <- read_tsv("data/metadata.tsv", show_col_types = T)
metafile$Sampleid <- NULL

# Taxonomy
rownames(taxa) <- NULL

# Create object
tse <- TreeSummarizedExperiment(assays = list(counts = counts),
                                rowData = DataFrame(taxa),
                                colData = DataFrame(metafile))

# Create novel variant names
names(tse) <- paste0("ASV", seq(nrow(tse)))

# Add representative sequences
repseq <- Biostrings::DNAStringSet(colnames(seqtab.nochim))
referenceSeq(tse) <- repseq 
```

\newpage

Our variant sequences are currently stored as names. They will moved to refseq and taxa names will be replaced by more convenient format

```{r, warning = FALSE, message = FALSE, size = "small"}
# Negative control sample 18 is unnecessary as number of counts are very low
tse <- tse[,colData(tse)$Name != "control"]
tse
# Keeping all taxa that are not unknown at Kingdom rank
tse <- tse[rowData(tse)$Kingdom == "Bacteria"]
tse
# Keeping all that are not Chloroplastic at Order rank
tse <- tse[rowData(tse)$Order != "Chloroplast" | is.na(rowData(tse)$Order)]
tse
# Keeping all that are not Mitochondrial at Family rank
tse <- tse[rowData(tse)$Family != "Mitochondria" | is.na(rowData(tse)$Family)]
tse
```

In the end we have 17 samples and 1525 taxa

\newpage

#### Write object data

Last step is to save data to suitable file formats.

All variant sequences are save to fasta

```{r, warning = F, message = F, size = "small"}

referenceSeq(tse) %>% writeXStringSet(paste0(exportloc,"/repseq.fasta"),
                                      append=F, compress=F, format="fasta")
```

Taxonomy table is converted to dataframe and written as tsv

```{r, warning = F, message = F, size = "small"}

taxonomy <- as.data.frame(rowData(tse)) %>% rownames_to_column(var="ASV")
write_tsv(taxonomy, paste0(exportloc, "/taxonomy.tsv"))
```

For metadata we add sampleid colum and write as tsv

```{r, warning = F, message = F, size = "small"}

metadf <- data.frame(sampleid=rownames(colData(tse)), colData(tse))
write_tsv(metadf, paste0(exportloc, "/meta.tsv"))
```

ASV count data need to be transposed prior writing

```{r, warning = F, message = F, size = "small"}

ASVdf <-  data.frame(ASV = rownames(tse),assay(tse, "counts"))
write_tsv(ASVdf, paste0(exportloc, "/asvs.tsv"))
```

TSE files is saved as RDS

```{r, warning = F, message = F, size = "small"}

saveRDS(tse, paste0(exportloc, "/tse.rds"))

```
